# Analysis of the steps activity

In this document, we will present you the 2 months of steps activity for an anonymous individual. Each measurement is made with an interval of 5 minutes which results of **17 568 observations**.

To reproduce the analysis you first need to get the data. They are accessible with the following link: [Activity.csv](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip).

Here the code to get the file and load the data.

```{r}
fileActivityCompressLink <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
fileActivityCompress <- tempfile()
download.file(fileActivityCompressLink, fileActivityCompress);
dataFileName <- "activity.csv"
fileActivityUncompress <- unzip(fileActivityCompress)
activityData <- read.csv(dataFileName)
head(activityData, 10)
```

## Analyses of the number of steps each day

In a first time, we will plot the total of steps done by the subject each day.

**(Warning: the package ggplot2 must be installed !)**

```{r, fig.height=4}
# install ggplot2 if not present for the generation of this document
if (!require("ggplot2",character.only = TRUE))
{
    install.packages("ggplot2",dep=TRUE)
}
library(ggplot2)
#clean data of na values
activityData <- activityData[!is.na(activityData$steps),]
# plot it
plt <- ggplot(data=activityData, aes(x=as.Date(date), y = steps)) + geom_bar(stat="identity") + ylab("Steps") + xlab("Dates")
print(plt)
```

In the second time, the median and the mean:

```{r}
# get the mean steps each days
meanActivityDataByDay <- tapply(activityData$steps, activityData$date, mean, na.rm = T, simplify = T)
# get the mean steps each days
medianActivityDataByDay <- tapply(activityData$steps, activityData$date, median, na.rm = T, simplify = T)
# group all the data
summaryActivityDataByDay <- data.frame(mean=meanActivityDataByDay, median=medianActivityDataByDay, date=names(meanActivityDataByDay), row.names = NULL)
# clean it
summaryActivityDataByDay <- summaryActivityDataByDay[!is.na(summaryActivityDataByDay$mea) | !is.na(summaryActivityDataByDay$med),]
#and display it
summaryActivityDataByDay
```

In a third time, we look for the steps average for each day:

```{r, fig.height=4}
plt <- ggplot(data=summaryActivityDataByDay, aes(x=as.Date(date), y = mean)) + geom_line() + ylab("mean steps") + xlab("Dates")
print(plt)
```

At last, We will observe the 5 minutes interval that, on average, contains the maximum of steps:

```{r}
# compute the maximun of step by day with a rankink system
# the ranking is invert with the min method and the setps are invert too.
# We this we have just to subset on the first rank to know the maximum
activityData$rank <- ave(-activityData$steps, activityData$date, FUN=function(x) rank(x, ties.method = "min"))
# find the maximum
maxStepByDay <- subset(activityData, rank == 1)
# found the average interval
meanMaxInterval <- mean(maxStepByDay$interval)
```
We found as an answer: `r meanMaxInterval`

# Handle missing values

To well understand the importance of missing values imputing strategy we will first measure the number of missing values present in the current dataset.

```{r, result="hide"}
activityData <- read.csv(dataFileName)
# simple method
nbrNaValue <- sum(length(which(is.na(activityData$steps))))
```

We found `r nbrNaValue` Na values in the dataset.

There is a lot of strategies to handle missing values. In the code above, we chose the strategies of simply discarding the value.

```{r, result="hide"}
# example
summaryActivityDataByDay <- summaryActivityDataByDay[!is.na(summaryActivityDataByDay$mea) | !is.na(summaryActivityDataByDay$med),]
``` 

But other strategies are possible. Few are described in the following document: [missing data imputation](http://www.stat.columbia.edu/~gelman/arm/missing.pdf).

For the rest of this document I chose to use the **Simple missing-data approaches that retain all the data**. This method does not change the size of the dataset. The missing values are simply replaced by a new one. Different startegies are used to choose the value. In this example, we will just replace it by **the mean for that 5 min interval**.

```{r}
activityDataMeanForInterval <- tapply(activityData$steps, activityData$interval, na.rm=T, mean)
for (i in 1:nrow(activityData))
{
 if (is.na(activityData[i,"steps"]))
    activityData[i,"steps"] <- activityDataMeanForInterval[as.character(activityData[i, "interval"])]
}
head(activityData)
```

With that strategy, we will be able to observe the impact of the imputation method on the final computation.

## Recompute the dataset

With this new missing value imputing strategies we recompute the total of steps done by the subject each day:

```{r, fig.height=4}
library(ggplot2)
#clean data of na values with the new method
activityData[is.na(activityData)] <- 0
# plot it
plt <- ggplot(data=activityData, aes(x=as.Date(date), y = steps)) + geom_bar(stat="identity") + ylab("Steps") + xlab("Dates")
print(plt)
```

And we recompute the mean and the median of those steps by days:

```{r}
# get the mean steps each days
meanActivityDataByDay <- tapply(activityData$steps, activityData$date, mean, na.rm = T, simplify = T)
# get the mean steps each days
medianActivityDataByDay <- tapply(activityData$steps, activityData$date, median, na.rm = T, simplify = T)
# group all the data
summaryActivityDataByDay <- data.frame(mean=meanActivityDataByDay, median=medianActivityDataByDay, date=names(meanActivityDataByDay), row.names = NULL)
# clean it
summaryActivityDataByDay <- summaryActivityDataByDay[!is.na(summaryActivityDataByDay$mea) | !is.na(summaryActivityDataByDay$med),]
#and display it
summaryActivityDataByDay
```

We can see that the results are really different. Particularly on the computation of the median and the mean.

- In the case of the sum of step number by day: the values by day are higher and some field which was empty are now well fill.
- In the case of the median/mean of the number of step by day: the median value which was constantly at 0 is now fill for some days.

## Comparing data

For finish, we will plot the difference between the average number of steps per 5 minutes interval across the weekdays and the weekends.

```{r, fig.height = 7, fig.width= 7}
if (!require("chron"))
{
  install.packages("chron")
}
library(chron)
par(mfrow=c(2, 1))

# found the appropriate day
activityData$daysweek <- sapply(as.Date(activityData$date), function(x) 
  if (is.weekend(x))
    "weekend"
  else
    "weekday"
    )
# get the data by days
activityDataByWeekDays <- split(activityData, activityData$daysweek)
# get day names
dayWeekNames <- unique(activityData$daysweek)
index <- 1
for (activityDataByWeekDay in activityDataByWeekDays)
{
  # get the data 
  activityDataWeekDayByInterval <- tapply(activityDataByWeekDay$steps, activityDataByWeekDay$interval, mean)
  plot(activityDataWeekDayByInterval, main=dayWeekNames[index], ylab = "Average steps", xlab = "Interval", type="l")
  index <- index + 1
}
```
